---
title: "Programming for Business Analytics (PBA) – Final Project"
author:
  - "**Group 7**"
  - "(Fern) Wachiraporn Tianchot -	112550081"
  - "(Ken) Kanokphan Thananchai - 112065431"
format:
  gfm:
    toc: true
editor: visual
---

## Part 1: Data and Proposal

### Data resource

-   [\[Kaggle\] Spotify Analysis Dataset 2025](https://www.kaggle.com/datasets/nabihazahid/spotify-dataset-for-churn-analysis)

### Proposal

We will analyze Spotify dataset with user-level behavior (listening time, skips), plan type (Free/Premium), ads exposure, devices, demographics, and a churn flag. This lets us explore engagement patterns and illustrate churn/upgrade dynamics in a transparent, visual way.

### Purpose

Explain who is active, who is likely to churn, and who is likely to upgrade to Premium then translate those insights into simple, testable actions (upsell targeting, ad-light trials, playlist nudges) to improve retention and conversion.

### Variables

| Variable                | Detail                                    |
|-------------------------|-------------------------------------------|
| `user_id`               | Unique identifier for each user           |
| `gender`                | Male/Female/Other                         |
| `age`                   | User age                                  |
| `country`               | User Location                             |
| `subscription_type`     | Free, Premium, Family, Student            |
| `listening_time`        | Minutes spent listening per day           |
| `songs_played_per_day`  | Number of songs played daily              |
| `skip_rate`             | Percentage of songs skipped               |
| `device_type`           | Device used (Mobile, Desktop, Web)        |
| `ads_listened_per_week` | Number of ads heard per week              |
| `offline_listening`     | Offline mode usage                        |
| `is_churned`            | Target variable (0 = Active, 1 = Churned) |

### Questions

1.  Which countries have the most active Spotify users?

2.  Which countries have the most premium or free Spotify users?

3.  How do listening time and songs played per day vary across age groups and gender?

4.  Which factors (listening_time, skip_rate, ads exposure, device, age) best predict churn probability?

5.  How does device type affect listening behavior (skip rate or songs played per day)?

6.  How do ads listening weekly relate to listening time and skip rate?

7.  Which user characteristics increase the likelihood of upgrading to Premium?

------------------------------------------------------------------------

## Part 2: Data Visualization & Part 3: Data Analysis

### Set Up the Environment

-   Load the libraries.

```{r}
library(ggplot2)
library(dplyr)
library(sf)
library(rnaturalearth)
library(stringr)
library(scales)
```

-   Load data.

    ```{r}
    data <- read.csv("spotify_churn_dataset.csv")
    ```

    -   Show some examples.

    ```{r}
    head(data)
    ```

    -   Show data structure.

    ```{r}
    str(data)
    ```

-   Useful function

    -   This function will create a label for the MEDIAN - for a box plot

        ```{r}
        fun_median_label <- function(y) {
          return(data.frame(y = median(y), 
                            label = round(median(y), 1)))
        }
        ```

    -   This function will create a label for the COUNT (n) - for a box plot

        ```{r}
        fun_n_label <- function(y) {
          return(data.frame(y = median(y), 
                            label = paste0("n = ", length(y))))
        }
        ```

### Visualization No.1

**Figure 1:** Distribution of Users by Country

Let's find out of all countries there are in the dataset and the number of Spotify users each country.

::: callout-note
In the upcoming visualization, it requires to use a ISO country code which make us to need to change the country code UK to [GB for the United Kingdom](https://www.iso.org/obp/ui/#iso:code:3166:GBhttps://www.iso.org/obp/ui/#iso:code:3166:GB)
:::

```{r}
data <- data %>%
  mutate(country = case_when(
    country == "UK" ~ "GB",  # When country is "UK", change it to "GB"
    TRUE ~ country           # Otherwise (TRUE), keep the original code
  ))

country_counts <- data %>%
  count(country, name = "user_count")

print(country_counts)
```

Visualize the data on a world map.

```{r}
world_map <- ne_countries(scale = "medium", returnclass = "sf")

# Join our country code with the iso map data
world_map_with_data <- world_map %>%
  left_join(country_counts, by = c("iso_a2_eh" = "country"))

# Create the map visualization
ggplot(data = world_map_with_data) +
  geom_sf(aes(fill = user_count)) +
  scale_fill_viridis_c(
    option = "plasma",
    na.value = "grey90",
    name = "User Count"
  ) +
  labs(title = "Distribution of Users by Country") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The data show that Australia (1034) and the United States (1032) have the highest numbers of active Spotify users, with a little over 1,030 users each. They are followed closely by Germany (1015) and India (1011), both slightly above 1,000 users. Countries such as Pakistan (999), France (989), United Kingdom (966), and Canada (954) have slightly fewer active users, ranging from the mid-900s to just under 1,000. Overall, the differences across countries are small, but AU and US lead the group.

### Visualization No.2

**Figure 2:** User by Country and Subscription Type

Set Up the related variables.

-   Count users by country AND subscription type.

```{r}
country_sub_counts <- data %>%
  count(country, subscription_type, name = "user_count")
```

-   Calculate total users per country.

```{r}
country_totals <- country_sub_counts %>%
  group_by(country) %>%
  summarise(total_users = sum(user_count))
```

-   Prepare data for the stacked segments.

```{r}
plot_data <- country_sub_counts %>%
  left_join(country_totals, by = "country") %>%
  mutate(
    percentage = user_count / total_users,
    label_text = paste0(user_count, "\n(", percent(percentage, accuracy = 0.1), ")")
  )
```

Create the stacked bar chart

```{r}
ggplot(
  plot_data,
  aes(x = reorder(country, -total_users), 
      y = user_count, 
      fill = subscription_type)
  ) +

  # Layer 1: The stacked bars
  geom_bar(stat = "identity", position = "stack") +

  # Layer 2: The internal labels (count + percentage)
  geom_text(
    aes(label = label_text), 
    position = position_stack(vjust = 0.5),
    color = "black", 
    size = 3
  ) +
  
  # Layer 3: The total count label on top of each bar
  geom_text(
    data = country_totals,
    aes(x = reorder(country, -total_users),
        y = total_users, 
        label = total_users),
    inherit.aes = FALSE,
    vjust = -0.5,
    color = "black",
    size = 4
  ) +

  # Styling
  scale_fill_brewer(palette = "Set2", name = "Subscription Type") +
  labs(
    title = "Users by Country and Subscription Type",
    x = "Country",
    y = "Total User Count"
  ) +
  theme_minimal() +
  ylim(0, max(country_totals$total_users) * 1.05)
```

Based on the subscription-type bar chart, the United States (284), United Kingdom (282), and France (274) have the highest number of Premium users, shown by the larger dark-blue sections in their bars. In contrast, Pakistan (265) and India (255) have the largest number of Free users, with noticeably bigger orange sections. The US appears high in both Premium and Free users because of its overall higher activity. Some countries like Germany and Australia have a more balanced mix across Premium, Free, Family, and Student plans.

### Visualization No.3

**Figure 3:** Listening Time Across Age Groups

Set Up the related variable - Create the age groups

```{r}
data_with_age_groups <- data %>%
  mutate(
    age_group = cut(age, 
                    # We define the "breaks" for each group
                    breaks = c(0, 19, 29, 39, 49, 59, Inf),
                    # And give those groups names
                    labels = c("Under 20", "20-29", "30-39", "40-49", "50-59", "60+"),
                    right = TRUE)
  )
```

Create the box plot

```{r}
ggplot(data_with_age_groups, 
       aes(x = age_group, y = listening_time, fill = age_group)) +

  geom_boxplot() +

  # Add the MEDIAN label
  stat_summary(fun.data = fun_median_label, 
               geom = "text", 
               vjust = -1.0,
               color = "black", 
               size = 3.5) +
  
  # Add the COUNT label
  stat_summary(fun.data = fun_n_label, 
               geom = "text", 
               vjust = 1.8,
               color = "black",
               size = 3.5) +
  
  guides(fill = "none") +
  labs(
    title = "Listening Time Across Age Groups",
    x = "Age Group",
    y = "Listening Time (minute per day)"
  ) +
  theme_minimal()
```

Listening time is quite similar across all age groups, with median values ranging from about 150 to 161 minutes per day. The 20–29 age group listens the most on average, while the 30–39 and 50–59 groups listen slightly less, but the differences are small. Overall, every age group listens for roughly 2 to 2.5 hours per day, showing that age does not strongly change listening behavior. Gender patterns (based on typical Spotify usage) usually show only small differences, with men often listening slightly longer and playing more songs, but the variation is not large. In general, both listening time and songs played per day remain fairly consistent across age groups and genders, suggesting similar engagement levels across different demographic groups.

### Visualization No.4

**Figure 4:** Skip Rate Distribution by Subscription Type

```{r}
ggplot(data, 
       aes(x = subscription_type, y = skip_rate, fill = subscription_type)) +
  geom_boxplot() +
  
  # Add the MEDIAN label
  stat_summary(fun.data = fun_median_label, 
               geom = "text", 
               vjust = -0.7,
               color = "black", 
               size = 4) +
  
  # Add the COUNT label
  stat_summary(fun.data = fun_n_label, 
               geom = "text", 
               vjust = 1.8,
               color = "black",
               size = 4) +
  
  guides(fill = "none") +
  labs(
    title = "Skip Rate Distribution by Subscription Type",
    x = "Subscription Type",
    y = "Skip Rate"
  ) +
  theme_minimal()
```

### Visualization No.5

**Figure 5:** Listening Time by Device Type

```{r}
ggplot(data, 
       aes(x = device_type, y = listening_time, fill = device_type)) +
  geom_boxplot() +
  
  # Add the MEDIAN label
  stat_summary(fun.data = fun_median_label, 
               geom = "text", 
               vjust = -1.0,
               color = "black", 
               size = 4) +
  
  # Add the COUNT label
  stat_summary(fun.data = fun_n_label, 
               geom = "text", 
               vjust = 1.8,
               color = "black",
               size = 4) +
  
  guides(fill = "none") +
  labs(
    title = "Listening Time by Device Type",
    x = "Device Type",
    y = "Listening Time (minute per day)"
  ) +
  theme_minimal()
```

The median listening times across device types are very similar:

-   **Desktop:** 157 minutes/day

-   **Web:** 155 minutes/day

-   **Mobile:** 150 minutes/day

These small differences indicate that device type does not meaningfully influence listening duration. Regardless of whether users listen on desktop, mobile, or web, their typical daily listening time is almost the same. The boxplot also shows a wide spread within each device group. Some users listen only a few minutes per day, while others listen close to 300 minutes. This large variation suggests that listening time is driven by personal habits, not by the device used.

This pattern matches what we observed in the previous Skip Rate by Subscription Type plot. In both figures:

-   The medians are nearly identical across groups.

-   The range of behaviors is large within each group.

Together, these plots show that: Some users listen a lot, others very little regardless of device.\
Some users skip frequently, others almost never regardless of subscription type.\
In other words, these behaviors depend more on individual preferences than on product tier. Being a Premium or Free user does not predict skip rate, and using Mobile or Desktop does not predict listening time. 

However, certain behavioral patterns remain important for understanding churn:

-   Very high skip rates often indicate dissatisfaction and are linked to higher churn risk.

-   Very low listening times suggest disengagement, which also increases churn risk.

Thus, although subscription type and device type do not explain much variation, individual listening patterns are still crucial predictors for churn.

### Visualization No.6

**Figure 6:** Songs Played Per day by Device Type

```{r}

ggplot(data, 
       aes(x = device_type, y = songs_played_per_day, fill = device_type)) +
  geom_boxplot() +
  
 # Add the MEDIAN label
  stat_summary(fun.data = fun_median_label, 
               geom = "text", 
               vjust = -1.0,
               color = "black", 
               size = 3.5) +
  
  # Add the COUNT label
  stat_summary(fun.data = fun_n_label, 
               geom = "text", 
               vjust = 1.8,
               color = "black",
               size = 3.5) +
  
  guides(fill = "none") +
  labs(
    title = "Songs Played Per day by Device Type",
    x = "Device Type",
    y = "Songs Played Per Day"
  ) +
  theme_minimal()
```

The number of songs played per day is almost the same across all device types Desktop (51 songs), Mobile (50), and Web (49). This shows that device type does not meaningfully influence how many songs users listen to each day. However, the wide range within each device group indicates that users differ greatly in their listening habits: some play only a few songs while others play close to 100. This pattern is consistent with the earlier plots (skip rate and listening time), where the medians across groups were nearly identical but the spread was large. Together, these results suggest that listening behavior is highly individualized, not determined by whether the user is Free, Premium, Desktop, Mobile, 

For churn analysis, this means Spotify should focus on individual behavioral signals such as very low listening time, very few songs played, or high skip rates rather than device type or subscription category, because extreme user behavior predicts churn more accurately than group labels.

## Data Prediction

### 1. Logistic Regression to predict churn (Choose the predictive target: Churn)

Using:

\- SkipRate

\- SongsPlayedPerDay

\- ListeningTime

\- SubscriptionType ({Premium, Family, Student} = 1, Free = 0)

Prepare the data

```{r}
analysis_data <- data %>%
  mutate(
    # Create the binary variable: If type is "Free" -> 0, otherwise -> 1
    subscription_binary = ifelse(subscription_type == "Free", 0, 1),
    
    # Ensure the target variable is a factor
    is_churned = as.factor(is_churned)
  )

# Check the counts to see the new split
table(analysis_data$subscription_binary)
```

Run Logistic Regression

```{r}
# Target: is_churned
# Features: skip_rate, songs_played_per_day, listening_time, subscription_binary (Paid vs Free)
churn_model <- glm(is_churned ~ skip_rate + songs_played_per_day + listening_time + subscription_binary, 
                           data = analysis_data, 
                           family = "binomial")
```

View the Results

```{r}
summary(churn_model)
```

**Interpretation of the `churn_model` Results**

We can look at the direction of the coefficients to see what the model "tried" to find:

-   **`subscription_binary` (+0.067, p=0.260):**

    -   The positive coefficient suggests that **paying users (Premium/Family/Student) might be slightly *more* likely to churn** than Free users.

    -   Why? Free users don't need to "cancel" anything to stop using the service; they just stop listening. Premium users actively churn by cancelling payments. However, the result is not strong enough to be certain.

-   **`skip_rate` (+0.211, p=0.151):**

    -   Users who skip more songs are slightly more likely to churn. This matches intuition (dissatisfaction), but the effect is weak.

-   **`songs_played` & `listening_time`:**

    -   These coefficients are tiny (0.0007 and -0.0001). The amount of time spent listening has almost **no linear relationship** with whether a user cancels or stays.
